\chapter{QEC decoders for color codes}
\label{chap:qec_decoders}

Having established the foundational principles of quantum error correction, and having explored both the structure and characteristics of color codes and the physical reality of error models, it is now time to address the final critical component of the QEC cycle: the decoders. A decoder is the classical algorithm responsible for transforming a set of syndrome measurements into a concrete diagnosis. In other words, it formulates a prediction of the error that has affected the logical observable, or equivalently, it infers the correct logical observable given the faulty measured one. The execution speed and accuracy of the decoders are major bottlenecks in realizing large-scale fault-tolerant quantum computation, and there's often a trade-off between them. Speed is relevant because the decoding process must ideally keep pace with the quantum circuit operation, as sometimes errors must be corrected during the execution. However, to increase decoding speed, accuracy may be sacrificed. 

\notatfm{The \textit{Decoder Error Model} is often referred to by different names in the literature: \textit{Syndrome Graph}, \textit{Decoding Graph} or \textit{Matching Graph}. Because our focus is in QEC simulation, we will prefer the term \textit{\textbf{Detector Error Model}} or DEM, which is the term used by \texttt{stim}. The DEM acts as a bridge between the physical noise phenomena (such as gate infidelities or readout errors) and the abstract graph or hypergraph structures used by most decoding algorithms to find the optimal correction.} To achieve high accuracy, a decoder must do more than just find any arbitrary error string that satisfies the observed syndrome; it must ideally identify the most probable error chain. At this point, we must introduce the concept of the \textit{Decoding Graph} or \textbf{Detector Error Model}, DEM. The DEM is a structured representation that captures the specific probabilities of the different error mechanisms that trigger detection events in the quantum circuit. By incorporating a DEM, decoders can leverage these probability distributions to make informed decisions. This additional statistical context allows the decoder to determine the actual error chain with a significantly higher success rate.

In this chapter, we will not dive into the programming aspects or the specific implementation details of decoders. The practical implementation of a basic decoder will be thoroughly covered in Chapter \ref{chap:implementation_and_simulation}, where it will also be compared against the standard Minimum Weight Perfect Matching (MWPM) algorithm to explicitly demonstrate the clear advantage of incorporating the DEM into the decoding process.


\section{Minimum Weight Perfect Matching (MWPM)}
\label{sec:mwpm}

\notatfm{MWPM relies in two fundamental things: \textbf{Perfect Matching}, meaning that it matches pairs of nodes with a set of edges, so no node is left unpaired and no node belongs to multiple matches, and \textbf{Minimum Weight}, meaning that the total weight of the selected set of edges is the minimum possible to perfectly match the active nodes.} The Minimum Weight Perfect Matching (MWPM) was the first algorithm to demonstrate that topological codes could be practically decoded. Its theoretical mapping to the surface code was established by Dennis, Kitaev, Landahl, and Preskill in 2002 \cite{dennis2002topological}. MWPM is arguably the most famous and widely used decoder in the field of quantum error correction. Originally conceived as a solution to a classical optimization problem by Jack Edmonds in 1965 \cite{edmonds1965paths}, MWPM is designed to find a perfect matching in a weighted graph such that the sum of the weights of the edges in the matching is minimized. 

In the context of QEC, and specifically for the Surface Code, the decoding problem maps naturally to a standard graph problem. When a single Pauli error occurs on a data qubit in a surface code, it typically flips the measurement outcome of exactly two adjacent stabilizers. These two activated stabilizers are the "defects" or "detection events" that form the syndrome. The decoding process consists of constructing a \textbf{syndrome graph} where:
\begin{itemize}
    \item \textbf{Nodes} represent the activated stabilizers.
    \item \textbf{Edges} represent the possible physical errors (or chains of errors) that could have caused those stabilizers to flip.
    \item \textbf{Weights} on the edges are assigned based on the error probabilities. Lower probability errors are assigned higher weights.
\end{itemize}




\begin{figure}[hbt!]
    \centering
    % SUBFIGURE A: PHYSICAL LATTICE
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}[scale=1.5]
            % Z-stabilizers (Red full faces)
            \foreach \x/\y in {0/0, 2/0, 1/1, 0/2, 2/2} {
                \filldraw[red!20, draw=red, thick] (\x,\y) rectangle (\x+1, \y+1);
                \node[font=\tiny, inner sep=0.5pt] at (\x+0.5, \y+0.62) {S(\x,\y)};
            }
            
            % Z-stabilizers (Red half faces - Rough boundaries)
            \filldraw[red!20, draw=red, thick] (0,1) -- (0,2) -- (-0.5, 1.5) -- cycle; 
            \node[font=\tiny, inner sep=0.5pt] at (-0.39, 1.5) {S(L,1)};
            
            \filldraw[red!20, draw=red, thick] (3,1) -- (3,2) -- (3.5, 1.5) -- cycle; 
            \node[font=\tiny, inner sep=0.5pt] at (3.39, 1.5) {S(R,1)};
        
            % X-stabilizers (Blue full faces)
            \foreach \x/\y in {1/0, 0/1, 2/1, 1/2} {
                \filldraw[blue!20, draw=blue, thick] (\x,\y) rectangle (\x+1, \y+1);
                \node[font=\tiny, inner sep=0.5pt] at (\x+0.5, \y+0.62) {S(\x,\y)};
            }
            
            % X-stabilizers (Blue half faces - Smooth boundaries)
            \foreach \x in {0, 2} {
                \filldraw[blue!20, draw=blue, thick] (\x,0) -- (\x+1,0) -- (\x+0.5, -0.5) -- cycle;
                \node[font=\tiny, inner sep=0.5pt] at (\x+0.5, -0.3) {S(\x,B)};
                
                \filldraw[blue!20, draw=blue, thick] (\x,3) -- (\x+1,3) -- (\x+0.5, 3.5) -- cycle;
                \node[font=\tiny, inner sep=0.5pt] at (\x+0.5, 3.3) {S(\x,T)};
            }
        
            % Ancilla Qubits (Solid black dots, reduced size)
            \foreach \x/\y in {0.5/0.5, 2.5/0.5, 1.5/1.5, 0.5/2.5, 2.5/2.5, 1.5/0.5, 0.5/1.5, 2.5/1.5, 1.5/2.5} {
                \filldraw[black] (\x,\y) circle (1.2pt);
            }
            \filldraw[black] (-0.16, 1.5) circle (1.2pt);
            \filldraw[black] (3.16, 1.5) circle (1.2pt);
            \foreach \x in {0.5, 2.5} {
                \filldraw[black] (\x, -0.16) circle (1.2pt);
                \filldraw[black] (\x, 3.16) circle (1.2pt);
            }
        
            % Data Qubits (White circles, reduced size)
            \foreach \x in {0, 1, 2, 3} {
                \foreach \y in {0, 1, 2, 3} {
                    \filldraw[black, fill=white, thick] (\x,\y) circle (1.5pt);
                    \node[above left, font=\tiny, inner sep=1pt] at (\x,\y) {D(\x,\y)};
                }
            }
            
        \end{tikzpicture}
        }
        \caption{}
        \label{fig:mwpm_lattice}
    \end{subfigure}
    \hfill
    % SUBFIGURE B: SYNDROME GRAPH
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}[scale=1.5]
            % Styles matching the new sizing and background removal
            \tikzset{
                syndrome/.style={circle, fill=black, inner sep=0pt, minimum size=5pt},
                defect/.style={circle, fill=red, inner sep=0pt, minimum size=5pt},
                edge/.style={gray!50, thick},
                edge_marker/.style={circle, draw=black, fill=white, thick, inner sep=0pt, minimum size=4pt},
                lbl/.style={font=\tiny, align=center, inner sep=1.5pt}
            }
            
            % Perfect 45-degree mathematical coordinates
            \coordinate (C00) at (0.5, 0.5);
            \coordinate (C20) at (2.5, 0.5);
            \coordinate (C11) at (1.5, 1.5);
            \coordinate (C02) at (0.5, 2.5);
            \coordinate (C22) at (2.5, 2.5);
            \coordinate (CHL) at (-0.5, 1.5); % Aligned to logical grid
            \coordinate (CHR) at (3.5, 1.5);  % Aligned to logical grid

            % 8 Boundary edges terminating at the virtual boundaries
            \foreach \start/\dest/\dlabel/\pos in {
                C00/{-0.5, -0.5}/{D(0,0)}/below right,
                C00/{1.5, -0.5}/{D(1,0)}/below left,
                C20/{1.5, -0.5}/{D(2,0)}/below right,
                C20/{3.5, -0.5}/{D(3,0)}/below left,
                C02/{-0.5, 3.5}/{D(0,3)}/above right,
                C02/{1.5, 3.5}/{D(1,3)}/above left,
                C22/{1.5, 3.5}/{D(2,3)}/above right,
                C22/{3.5, 3.5}/{D(3,3)}/above left%
            } {
                \draw[edge] (\start) -- (\dest) node[midway, edge_marker] {} node[midway, \pos, lbl] {\dlabel\\$w=1$};
            }

            % Path 2 (Sub-optimal, blue dashed)
            \foreach \start/\dest/\dlabel/\weight/\pos in {
                C00/CHL/{D(0,1)}/1/above right,
                CHL/C02/{D(0,2)}/1/above left,
                C02/C11/{D(1,2)}/1/above right,
                C11/C22/{D(2,2)}/2/above left,
                C22/CHR/{D(3,2)}/2/above right%
            } {
                \draw[blue, dashed, very thick] (\start) -- (\dest) node[midway, solid, edge_marker] {} node[midway, \pos, lbl, text=blue] {\dlabel\\$w=\weight$};
            }

            % Path 1 (Optimal, red solid)
            \foreach \start/\dest/\dlabel/\weight/\pos in {
                C00/C11/{D(1,1)}/1/below right,
                C11/C20/{D(2,1)}/1/below left,
                C20/CHR/{D(3,1)}/1/below right%
            } {
                \draw[red, ultra thick] (\start) -- (\dest) node[midway, edge_marker] {} node[midway, \pos, lbl, text=red] {\dlabel\\$w=\weight$};
            }

            % Place the 7 graph nodes covering the ends of the lines
            \node[syndrome] at (C20) {}; \node[below=2pt, font=\tiny] at (C20) {S(2,0)};
            \node[syndrome] at (C11) {}; \node[above=2pt, font=\tiny] at (C11) {S(1,1)};
            \node[syndrome] at (C02) {}; \node[above=2pt, font=\tiny] at (C02) {S(0,2)};
            \node[syndrome] at (C22) {}; \node[above=2pt, font=\tiny] at (C22) {S(2,2)};
            \node[syndrome] at (CHL) {}; \node[left=2pt, font=\tiny] at (CHL) {S(L,1)};
            
            % Highlight the active defects in red
            \node[defect] at (C00) {}; \node[below left=1pt, font=\tiny, text=red] at (C00) {S(0,0)};
            \node[defect] at (CHR) {}; \node[right=2pt, font=\tiny, text=red] at (CHR) {S(R,1)};

            % Display total path weights
            \node[blue, font=\small, align=center] at (1.5, 3.8) {Sub-optimal path\\Total $w=7$};
            \node[red, font=\small, align=center] at (2.0, -0.8) {Minimum weight path\\Total $w=3$};
            
        \end{tikzpicture}
        }
        \caption{}
        \label{fig:mwpm_graph}
    \end{subfigure}
    
    \caption{A visual representation of the Minimum Weight Perfect Matching problem under a code-capacity noise model. \textbf{(a)} The rotated surface code lattice introduced in Chapter~\ref{chap:state_of_the_art}. Data qubits are labeled as $D(x,y)$ and ancillas (stabilizers) as $S(x,y)$. \textbf{(b)} The full syndrome matching graph specifically for X-type errors, consisting of 7 nodes (the Z-stabilizer ancillas, shown as small black circles) and 16 edges (one for each physical data qubit--each possible X-type error in the code-capacity model--, represented by white circles indicating the weight (likeliness) of this error. Two possible error chains are highlighted, one of them being the one with minimum weight.}
    \label{fig:4:mwpm_example}
\end{figure}

The MWPM algorithm efficiently searches this graph to find the pairings of defects that minimize the total weight. This is illustrated in Figure \ref{fig:mwpm_graph}. Once the optimal pairing is found, the decoder identifies the most probable error chain and applies the corresponding correction. This approach is highly effective and its implementation (\cite{higgott2022pymatching}) serves as the benchmark against which almost all new decoders are measured.

\subsection{Edmonds' Blossom algorithm}
To understand how MWPM achieves this globally optimal pairing, it is necessary to look at its underlying mathematical engine: Edmonds' Blossom algorithm. A naive greedy approach—simply pairing the closest defects as soon as they are found—can easily get trapped in local minima, leaving isolated defects that force very high-weight connections later. Edmonds' algorithm systematically explores the graph to guarantee the absolute minimum weight solution.

This exploration is intuitively visualized using expanding regions. The algorithm begins by simultaneously growing a "bubble" outwards from every active defect. These bubbles expand across the adjacent data qubits in the graph, to other non-active defects. When the boundaries of two expanding bubbles collide, the two defects are temporarily matched. The algorithm assumes that the path followed by the bubbles corresponds to the chain of physical errors that connected them.

\notatfm{Let's see an example. Defects $A$ and $B$ are matched ($A \Rightarrow B$) and then defect $C$'s expanding bubble reaches $B$.  The path between them is stored as a non-matched path ($C \rightarrow B$). Defect $A$ is re-activated and it's bubble begins growing again. If the bubble reaches an active unmatched defect $D$, then an alternating path forms ($D \rightarrow A \Rightarrow B \rightarrow C$). This path has two non-matched segments and one matched segment. The algorithm "flips" the entire path's status ($D \Rightarrow A \rightarrow B \Rightarrow C$), correcting past suboptimal assumptions and increasing the total number of matched pairs.}As the bubbles continue to grow, they often encounter defects that are already matched to someone else, creating topological conflicts. The algorithm efficiently resolves these conflicts by re-evaluating its temporary matchings as new connections are discovered. Conflicting matches are resolved using \textit{alternating paths}—routes that alternate between currently assumed errors (matched edges) and assumed healthy qubits (unmatched edges). If an alternating path is found connecting two free defects, the algorithm "flips" the entire path's status, simultaneously correcting past suboptimal assumptions and increasing the total number of matched pairs.

Edmonds' algorithm succeeds where other algorithms fail because of how it handles cycling paths: the temporary connections form a closed loop consisting of an odd number of nodes. Odd-length cycles can't be paired internally (the flip will not fix the alternating chain). The Blossom algorithm overcomes this by temporarily collapsing the entire odd-cycle into a single "super-node" or \textit{blossom}. This new blossom then resumes expanding as a single entity to find an external pair. Once all defects in the graph are paired, the algorithm unfolds the blossoms back into their original nodes and resolves the internal pairings. This continuous process of growing bubbles, shrinking odd-cycles, and unfolding them guarantees finding the most probable physical error chain.

\subsubsection*{The Challenge of MWPM in Color Codes}

While MWPM is exceptionally well-suited for Surface Codes, it cannot be directly applied to 2D Color Codes. The reason lies in the fundamental topology of the code.

In a surface code, a physical error creates a 1-dimensional string connecting two defects. However, in a color code, a single physical Pauli error activates \textbf{three} adjacent stabilizers. Consequently, an error chain in a color code is not a simple path between two points, but a branching, 2-dimensional structure that connects multiple groups of three defects.

In the following sections we will study different alternatives to decode color codes, some of which heavily rely on MWPM.

\section{Union-find}
\label{sec:union_find}

\notatfm{The overall worst-case complexity is $O(n \alpha(n))$, where $n$ is the number of physical qubits and $\alpha(n)$ is the inverse Ackermann function—a function that grows so slowly that it practically evaluates to less than 3.} While the Minimum Weight Perfect Matching (MWPM) algorithm sets the standard for decoding accuracy, its worst-case computational complexity does not scale well with the number of nodes in the syndrome graph. Decoders must operate fast enough to prevent a backlog of uncorrected errors during the circuit execution, wich causes a bottleneck as fault-tolerant architectures scale to thousands of physical qubits. To address this speed bottleneck, Delfosse and Nickerson introduced the Union-find decoder in 2018 \cite{Delfosse_2021}, achieving an almost-linear worst-case complexity while maintaining a decoding threshold highly comparable to MWPM.

The primary innovation of the Union-find decoder is the replacement of the global path optimization of Edmonds' Blossom algorithm with a greedy solution. As mentioned earlier, a naive greedy approach would easily fall into local minima. Union-find circumvents this by proposing a greedy cluster-growth strategy instead. Similar to MWPM, bubbles grow and merge upon collision, but rather than evaluating the optimal pairing when two bubbles collide, the algorithm simply continues growing them as long as they contain an odd number of defects. Once every cluster contains an even number of defects (meaning all defects are safely encapsulated), a secondary phase acts strictly within each cluster to definitively pair the defects. The algorithm thus operates in two strictly sequential phases, as seen in Figure \ref{fig:union_find_steps}:
\begin{enumerate}
    \item \notatfm{The \textbf{Disjoint-Set} data structure tracks elements partitioned into disjoint subsets, supporting two efficient operations: \textbf{Find} identifies the root of a subset (the unique element of the subset that identifies this subset), and \textbf{Union} merges two subsets. Checking or merging clusters is executed in practically constant time, bringing the computational cost of the algorithm to near-linear with respect to the number of nodes.} \textbf{Cluster Growth:} Initially, every individual defect is placed in its own separate cluster. Using the standard \textit{Disjoint-Set} (or Union-find) data structure, the algorithm tracks these disjoint clusters. A cluster is classified as \textit{valid} if it contains an even number of defects, or if it touches a boundary. If a cluster is valid, its growth pauses. Conversely, any \textit{invalid} (odd-parity) cluster grows uniformly by absorbing its neighboring edges. When two growing clusters collide, the algorithm executes a \texttt{union} operation to merge them into a single, larger cluster. This iterative growth and merging process continues until every cluster in the graph becomes valid. As opposed to MWPM's blossom contraction, the Disjoint-Set data structure allows an easier handling of a cluster colliding with itself (creating an internal loop) thanks to the \textit{find} operator.
    
    \item \notatfm{At an intersection, if both (or neither) of the child branches bring an unpaired defect, the error strings are combined and the upwards node is left unmarked. If only one of the two branches brings an unpaired defect, then the upwards node is marked as an unpaired defect so the pruning can continue upwards.} \textbf{Path Resolution:} Once all clusters are validated, the algorithm must decide which specific edges inside each cluster correspond to the physical errors. For each valid cluster, the Union-Find process has already defined a tree (i.e. a graph without loops). This allows applying a highly efficient \textit{peeling algorithm}: starting from the leaf nodes of the tree and working toward the root, the algorithm greedily assigns error correction strings to pair all internal defects. Each leaf node is evaluated and then pruned. If the node is a defect, the pruned edge is added to an error correction string and the next upwards node is marked as an unpaired defect to continue the pruning from it. When two branches intersect, the pruning on that path pauses until both branches reach the intersection point.
\end{enumerate}

\begin{figure}[hbt!]
    \centering
    
    % Definición rigurosa de ejes activos por iteración para cadena de 4 errores
    \newcommand{\rzero}{(6,2)--(7,2) (6,2)--(6,3)}
    
    \newcommand{\lone}{(2,2)--(1,2) (2,2)--(3,2) (2,2)--(2,1) (2,2)--(2,3)}
    
    \newcommand{\rone}{(5,2)--(6,2) (6,1)--(6,2) (7,2)--(8,2) (7,2)--(7,1) (7,2)--(7,3) (6,3)--(5,3) (6,3)--(6,4) (6,3)--(7,3)}
    
    \newcommand{\ltwo}{(0,2)--(1,2) (1,1)--(1,2) (1,3)--(1,2) (4,2)--(3,2) (3,1)--(3,2) (3,3)--(3,2) (2,0)--(2,1) (1,1)--(2,1) (3,1)--(2,1) (2,4)--(2,3) (1,3)--(2,3) (3,3)--(2,3)}
    
    \newcommand{\rtwo}{(4,2)--(5,2) (5,1)--(5,2) (5,2)--(5,3) (5,1)--(6,1) (7,1)--(6,1) (6,0)--(6,1) (9,2)--(8,2) (8,1)--(8,2) (8,3)--(8,2) (7,0)--(7,1) (8,1)--(7,1) (8,3)--(7,3) (7,4)--(7,3) (4,3)--(5,3) (5,4)--(5,3) (5,4)--(6,4) (7,4)--(6,4) (6,5)--(6,4)}
    
    \tikzset{
        syndrome/.style={circle, fill=black, inner sep=0pt, minimum size=2.5pt},
        defect/.style={circle, fill=red, inner sep=0pt, minimum size=5pt},
        edge/.style={gray!30, very thin},
        error/.style={red, ultra thick},
        erasure/.style={orange, ultra thick, dashed},
        activeedge/.style={blue, ultra thick},
        envleft/.style={magenta!10, line width=18pt, line cap=round, line join=round},
        envright/.style={magenta!25, line width=18pt, line cap=round, line join=round},
        envvalid/.style={green!25, line width=18pt, line cap=round, line join=round}
    }
    
    % (a) Initial
    \begin{subfigure}[b]{0.19\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \foreach \x in {0,...,9} { \draw[edge] (\x,0) -- (\x,5); }
            \foreach \y in {0,...,5} { \draw[edge] (0,\y) -- (9,\y); }
            
            \draw[erasure] \rzero;
            % Cadena de error de longitud 4
            \draw[error] (2,2) -- (3,2) -- (4,2) -- (5,2) -- (6,2);
            
            \foreach \x in {0,...,9} { \foreach \y in {0,...,5} { \node[syndrome] at (\x,\y) {}; } }
        \end{tikzpicture}
        }
        \caption{Initial}
    \end{subfigure}
    \hfill
    % (b) Syndrome (t=0)
    \begin{subfigure}[b]{0.19\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \fill[envleft] (2,2) circle (9pt);
            \draw[envright] \rzero;
            
            \foreach \x in {0,...,9} { \draw[edge] (\x,0) -- (\x,5); }
            \foreach \y in {0,...,5} { \draw[edge] (0,\y) -- (9,\y); }
            
            \draw[activeedge] \rzero;
            
            \foreach \x in {0,...,9} { \foreach \y in {0,...,5} { \node[syndrome] at (\x,\y) {}; } }
            \node[defect] at (2,2) {}; \node[defect] at (6,2) {};
        \end{tikzpicture}
        }
        \caption{Syndrome}
    \end{subfigure}
    \hfill
    % (c) Growth 1 (t=1)
    \begin{subfigure}[b]{0.19\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \draw[envleft] \lone;
            \draw[envright] \rzero \rone;
            
            \foreach \x in {0,...,9} { \draw[edge] (\x,0) -- (\x,5); }
            \foreach \y in {0,...,5} { \draw[edge] (0,\y) -- (9,\y); }
            
            \draw[activeedge] \lone;
            \draw[activeedge] \rzero \rone;
            
            \foreach \x in {0,...,9} { \foreach \y in {0,...,5} { \node[syndrome] at (\x,\y) {}; } }
            \node[defect] at (2,2) {}; \node[defect] at (6,2) {};
        \end{tikzpicture}
        }
        \caption{Growth}
    \end{subfigure}
    \hfill
    % (d) Growth 2 (t=2) - COLLISION
    \begin{subfigure}[b]{0.19\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            % Se dibujan en sus propios colores, solapando exactamente en el nodo (4,2)
            \draw[envleft] \lone \ltwo;
            \draw[envright] \rzero \rone \rtwo;
            
            \foreach \x in {0,...,9} { \draw[edge] (\x,0) -- (\x,5); }
            \foreach \y in {0,...,5} { \draw[edge] (0,\y) -- (9,\y); }
            
            \draw[activeedge] \lone \ltwo;
            \draw[activeedge] \rzero \rone \rtwo;
            
            \foreach \x in {0,...,9} { \foreach \y in {0,...,5} { \node[syndrome] at (\x,\y) {}; } }
            \node[defect] at (2,2) {}; \node[defect] at (6,2) {};
        \end{tikzpicture}
        }
        \caption{Growth}
    \end{subfigure}
    \hfill
    % (e) Merge (t=2 confirmed)
    \begin{subfigure}[b]{0.19\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            % El envoltorio ahora es verde para todo el clúster unificado
            \draw[envvalid] \lone \ltwo \rzero \rone \rtwo;
            
            \foreach \x in {0,...,9} { \draw[edge] (\x,0) -- (\x,5); }
            \foreach \y in {0,...,5} { \draw[edge] (0,\y) -- (9,\y); }
            
            \draw[activeedge] \lone \ltwo \rzero \rone \rtwo;
            
            \foreach \x in {0,...,9} { \foreach \y in {0,...,5} { \node[syndrome] at (\x,\y) {}; } }
            \node[defect] at (2,2) {}; \node[defect] at (6,2) {};
        \end{tikzpicture}
        }
        \caption{Merge (Valid)}
    \end{subfigure}

    \caption{Visual representation of the Union-find decoder cluster-growth strategy: edges are data qubits and vertices are the detectors connected to these data qubits. \textbf{(a)} An initial error chain (red) occurs on the lattice, alongside two known erased data qubits (dashed orange), meaning there's prior knowledge of these two data qubits being affected by loss or leakage. \textbf{(b)} The decoder identifies the syndrome defects (red dots) and defines the two odd-parity (invalid) clusters (light and dark magenta). The erased qubits are incorporated to the right cluster. \textbf{(c)} The algorithm grows the clusters uniformly. Because of the initial erasures, the right cluster expands faster. \textbf{(d)} In the second growth step, the boundaries of both clusters collide at a shared node. \textbf{(e)} This collision immediately triggers a \texttt{union} operation. The newly merged cluster now contains an even number of defects, changing its overall status to \textit{valid} (green) and halting the growth phase.}
    \label{fig:union_find_steps}
\end{figure}



The algorithm described so far does not account for different error probabilities. Because of that, it can lead to lower accuracy than MWPM. \textit{Weighted} variants of Union-find incorporate the DEM edge weights directly into the growth phase, causing clusters to grow faster through edges with higher probabilities (lower weights). This bridges the accuracy gap with MWPM while retaining fast execution speeds.

\subsubsection*{Application to Color Codes}
\notatfm{Note that the Union-Find algorithm has mechanisms to deal with erasure (known loss or leakage of a specific qubit). In the initial state, the edges corresponding to these qubits are accounted for when determining the initial clusters. In the simple code-capacity model that we will follow in this work, the initial erasure will always be empty.} Like MWPM, the standard Union-find decoder is fundamentally a graph-based algorithm designed to pair defects connected by 1-dimensional strings. Consequently, applying Union-find directly to the branching error chains (string-nets) of 2D Color Codes presents the exact same topological challenge outlined in Section~\ref{sec:mwpm}. To decode color codes efficiently, Union-find is typically integrated as the fast computational engine within reduction frameworks, such as Projection or Restriction decoders, which map the complex hypergraph onto standard surface code sub-graphs.

\section{BP+OSD} 
\label{sec:bp_osd}

\notatfm{In classical and quantum coding theory, a \textbf{Tanner graph} is a bipartite graph used to represent the parity-check matrix of a code. It consists of two sets of vertices: \textit{variable nodes} (representing the physical data qubits) and \textit{check nodes} (representing the stabilizers). An edge connects a variable node to a check node if that qubit is involved in that stabilizer measurement.}While Minimum Weight Perfect Matching and Union-find are highly effective, they are fundamentally graph-based algorithms designed for 1-dimensional errors (strings). As discussed, directly applying them to the branching error chains (string-nets) of 2D Color Codes is not naturally possible. Belief Propagation with Ordered Statistic Decoding (BP+OSD) represents a complete paradigm shift: rather than trying to solve a geometrical routing problem on a lattice, BP+OSD treats decoding as a purely algebraic problem based on the parity-check matrix of the code. This makes it natively compatible with any quantum code, including color codes, without requiring any complex geometrical mappings or projections.

The BP+OSD decoder consists of two algorithms working sequentially. First, Belief Propagation leverages the Tanner graph, iteratively evaluating the syndrome to estimate the error probabilities of the individual qubits. Then, if this process fails to converge to a valid configuration, Ordered Statistic Decoding serves as a post-processing step, using the previously calculated probabilities to enforce a valid mathematical solution.

\subsubsection*{Belief Propagation (BP)}

\notatfm{Unlike MWPM or Union-find, which only use \textit{hard information} (binary 0 or 1 from the syndrome), Belief Propagation utilizes \textit{soft information}, continuously updating a continuous probability for each qubit. Mathematically, this relies on the Sum-Product Algorithm: a check node deduces the error probability of a target qubit by using Gallager's formula \cite{Gallager_1962_ldpc} to compute the likelihood that all its \textit{other} connected qubits possess a collective parity that matches the measured syndrome. Note that to calculate the error probability of a certain variable qubit, the algorithm only uses extrinsic information (the information from all the other variable qubits and the meassured syndrome).}Belief Propagation is a classical message-passing algorithm proposed by Pearl in 1988 \cite{Pearl_88}. It operates on the \textit{Tanner graph} of the code. BP works by having nodes "talk" to each other iteratively. 

Initially, the \textit{DEM} information is used to assign error probabilities to the variable nodes (data qubits), which then send this probability to the check nodes. 

The check nodes process these newly received probabilities to enforce the code's parity rules. To update a particular variable node, a check node looks at its own syndrome outcome and the estimated error probabilities of all the other variable nodes attached to it. From this collective information, it deduces the likelihood that the target variable node has an error, and sends this newly refined probability back to it. 

The variable nodes gather these messages from all the stabilizers they belong to, update their own internal probability of holding an error, and send a message back to the check nodes, informing them of their evaluation of their own error probability.

This continuous back-and-forth exchange, represented in Figure \ref{fig:bp_convergence}, iteratively refines the global distribution of errors. This cycle continues until the variable nodes' probabilities converge to a clear, definitive error pattern that perfectly satisfies the observed syndrome. 

\begin{figure}[hbt!]
    \centering
    
    \tikzset{
        var/.style={circle, draw=black, thick, fill=white, minimum size=20pt, inner sep=0pt, font=\small},
        checkactive/.style={rectangle, draw=red!80!black, thick, fill=red!20, minimum size=18pt, inner sep=0pt, font=\small},
        checkinactive/.style={rectangle, draw=black, thick, fill=gray!20, minimum size=18pt, inner sep=0pt, font=\small},
        edge/.style={gray!70, thick},
        msgup/.style={->, >=stealth, blue, thick, shorten >= 8pt, shorten <= 8pt},
        msgdown/.style={->, >=stealth, green!60!black, thick, shorten >= 8pt, shorten <= 8pt},
        lbl/.style={fill=white, inner sep=1pt, font=\scriptsize, text=black}
    }
    
    \newcommand{\drawnodes}{
        \node[var] (V1) at (0.5,0) {$v_1$};
        \node[var] (V2) at (2.5,0) {$v_2$};
        \node[var] (V3) at (4.5,0) {$v_3$};
        \node[var] (V4) at (6.5,0) {$v_4$};
        
        \node[checkactive] (C1) at (1.5, 2.5) {$c_1$};
        \node[checkactive] (C2) at (3.5, 2.5) {$c_2$};
        \node[checkinactive] (C3) at (5.5, 2.5) {$c_3$};
    }
    
    \newcommand{\drawedges}{
        \draw[edge] (V1) -- (C1);
        \draw[edge] (V2) -- (C1);
        \draw[edge] (V2) -- (C2);
        \draw[edge] (V3) -- (C2);
        \draw[edge] (V3) -- (C3);
        \draw[edge] (V4) -- (C3);
    }
    
    % (a) Initialization
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \drawnodes
            \drawedges
            
            \node[below=8pt, font=\small] at (V1) {$p=0.1$};
            \node[below=8pt, font=\small] at (V2) {$p=0.4$};
            \node[below=8pt, font=\small] at (V3) {$p=0.1$};
            \node[below=8pt, font=\small] at (V4) {$p=0.1$};
        \end{tikzpicture}
        }
        \caption{Initialization}
    \end{subfigure}
    \hfill
    % (b) Variable to Check
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \drawnodes
            \drawedges
            
            \draw[msgup, bend left=15] (V1) to node[lbl] {0.1} (C1);
            \draw[msgup, bend right=15] (V2) to node[lbl] {0.4} (C1);
            
            \draw[msgup, bend left=15] (V2) to node[lbl] {0.4} (C2);
            \draw[msgup, bend right=15] (V3) to node[lbl] {0.1} (C2);
            
            \draw[msgup, bend left=15] (V3) to node[lbl] {0.1} (C3);
            \draw[msgup, bend right=15] (V4) to node[lbl] {0.1} (C3);
            
            \node[below=8pt, font=\small] at (V1) {$p=0.1$};
            \node[below=8pt, font=\small] at (V2) {$p=0.4$};
            \node[below=8pt, font=\small] at (V3) {$p=0.1$};
            \node[below=8pt, font=\small] at (V4) {$p=0.1$};
        \end{tikzpicture}
        }
        \caption{Variable to Check}
    \end{subfigure}
    
    \vspace{0.6cm} % Separación entre filas
    
    % (c) Check to Variable
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \drawnodes
            \drawedges
            
            \draw[msgdown, bend left=15] (C1) to node[lbl] {0.6} (V1);
            \draw[msgdown, bend right=15] (C1) to node[lbl] {0.9} (V2);
            
            \draw[msgdown, bend left=15] (C2) to node[lbl] {0.9} (V2);
            \draw[msgdown, bend right=15] (C2) to node[lbl] {0.6} (V3);
            
            \draw[msgdown, bend left=15] (C3) to node[lbl] {0.1} (V3);
            \draw[msgdown, bend right=15] (C3) to node[lbl] {0.1} (V4);
            
            \node[below=8pt, font=\small] at (V1) {$p=0.1$};
            \node[below=8pt, font=\small] at (V2) {$p=0.4$};
            \node[below=8pt, font=\small] at (V3) {$p=0.1$};
            \node[below=8pt, font=\small] at (V4) {$p=0.1$};
        \end{tikzpicture}
        }
        \caption{Check to Variable}
    \end{subfigure}
    \hfill
    % (d) Convergence
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
        \begin{tikzpicture}
            \drawnodes
            \drawedges
            
            % Highlight del nodo convergido
            \node[var, fill=green!20] at (2.5,0) {$v_2$};
            
            \node[below=8pt, font=\small] at (V1) {$p=0.14$};
            \node[below=8pt, font=\small] at (V2) {$p=0.98$};
            \node[below=8pt, font=\small] at (V3) {$p=0.02$};
            \node[below=8pt, font=\small] at (V4) {$p=0.01$};
        \end{tikzpicture}
        }
        \caption{Convergence}
    \end{subfigure}

    \caption{Visual representation of the Belief Propagation (BP) algorithm numerical updates on a simple Tanner graph. \textbf{(a)} Variables (circles) are initialized with prior probabilities $p$ based on the \textit{DEM}. Checks (squares) observe the syndrome; active syndromes are marked in red, while inactive ones are gray. \textbf{(b)} Variable-to-Check step: Variables broadcast their current probability to connected checks (blue arrows). \textbf{(c)} Check-to-Variable step: Checks enforce parity rules. For example, the active check $c_1$ requires odd parity; knowing $v_1$ has only a $10\%$ chance of error, it deduces that $v_2$ must have a $90\%$ chance of being the error, sending this refined extrinsic probability back (green arrows). Conversely, the inactive check $c_3$ requires even parity; assuming $v_4$ is healthy ($10\%$), it deduces $v_3$ must also be healthy ($10\%$). \textbf{(d)} Convergence: Variables aggregate the incoming evidence using Bayes' rule. Node $v_2$ receives strong confirmation from both active checks, spiking its probability to $0.98$. Node $v_3$ is heavily suppressed to $0.02$ because the inactive check $c_3$ vouches for its health. The algorithm flawlessly converges to the true physical error.}
    \label{fig:bp_convergence}
\end{figure}


If the Tanner graph has a tree-like structure, BP is mathematically guaranteed to find the exact optimal solution. However, in topological codes like the Color Code, adjacent stabilizers share multiple data qubits. This translates into a Tanner graph with short, closed loops. In a graph with many short loops, the messages passed by the nodes start echoing in circles. A node might receive a high probability of error simply because its own previous message traveled through a loop and came back amplified. This phenomenon creates \textit{trapping sets}, where the probabilities wildly oscillate and the BP algorithm fails to converge to a valid physical error chain, rendering it unusable on its own for surface or color codes.

\subsubsection*{Ordered Statistic Decoding (OSD)}

\notatfm{The algebraic nature of OSD involves Gaussian elimination over the parity-check matrix. In simple terms, the algorithm locks in the assumed state of the qubits that BP is most certain about. By substituting these "reliable" qubits into the system of equations that defines the parities as a function of the qubit errors, it step-by-step isolates the remaining uncertain qubits to logically force a valid error pattern. However, chaining these mathematical deductions across the entire code has a worst-case computational complexity of $O(N^3)$. While this makes BP+OSD remarkably accurate for color codes, it also makes it significantly slower than Union-find for large distances, presenting a classical trade-off between speed and decoding threshold.} To overcome BP's failure to converge, Roffe introduced Ordered Statistic Decoding (OSD) as a post-processing step \cite{Roffe_2020}. When BP gets trapped and fails to find a valid solution, it doesn't mean the algorithm was useless. Even if the probabilities oscillated, the final "soft" probabilities calculated by BP still contain incredibly valuable information about which qubits are most likely to be corrupted.

OSD steps in by taking these final, unconverged probabilities from BP and sorting all the data qubits from the most likely to have an error to the least likely. It then selects a subset of the most strongly suspected failing qubits. Using standard linear algebra--specifically, Gaussian elimination over the code's parity-check matrix--OSD forces a valid mathematical solution that aligns with the syndrome.

By combining the probabilistic insight of Belief Propagation with the brute-force algebraic resolution of Ordered Statistic Decoding, BP+OSD bypasses the hypergraph problem of Color Codes and achieves some of the highest decoding accuracy thresholds.

\section{Projection decoders (Delfosse, 2014)}

\section{Restriction decoders (Kubica \& Delfosse, 2023)}

\section{Möbius decoder (Sahay \& Brown, PRX Quantum, 2022)}

\section{Concatenated MWPM decoder (Lee, Li \& Bartlett, Quantum, 2025)}

\section{Correlated matching decoder (Liu, Wu \& Lao, 2025)}
