{\LARGE\bfseries\color{nebrijared}Acknowledgments\par}
\fontsize{11}{13}\selectfont

\begin{itshape}
I would like to express my deepest gratitude to everyone who made this Master's Thesis possible. 

First and foremost, I want to thank my parents, Frasco and María José, who instilled in me freedom, curiosity, and the courage to challenge the status-quo. They allowed me to grow free from the weight of expectations and always appreciated and celebrated my achievements. My gratitude also goes to my godfather, Nono, for teaching me perseverance, a self-respect that strives for excellence, and the profound desire to leave the world a better place than I found it; and to my grandparents, who taught me the true value of knowledge, wisdom, self-denial, and hard work. And last, but not least among my roots, I want to thank Lolo, who taught me that there are many different, yet equally valid, ways to approach life, and that the journey is more important than the destination.

My wife, Nines, and our children, Pedro, Antonio, and Carlos also deserve all my gratitude. They are not only my primary motivation, but they have also sacrificed as much as I have for this project. Pursuing a Master's degree in one's forties, and doing it while working full-time, demands immense family sacrifices, and I could not have reached this finish line without their patience, understanding, and love.

I owe much to my professors in this Master's program, and by extension to all my mentors and teachers, who have built the foundation that has allowed me to reach this point: Those who teach us are the invisible architects of our identity. Among them, I want to especially thank Sergio Boixo; his talk first inspired me to delve into the world of quantum computing, and his responses when I asked for advice helped me take my first steps in this field. I also owe a great debt of gratitude to Greg Quillard, a good friend who happens to be the person who has fostered my professional growth the most throughout my career.

Last, but certainly most important, I would like to express my sincere appreciation to the scientific advisor of this work, Miguel Ángel. It has been an absolute privilege to have this thesis directed by such an eminence in the field, and it is a unique honor to write about 2D color codes under the guidance of their co-creator. I am deeply grateful for his time, his insight, and this extraordinary opportunity. Additionally, I extend my thanks to my academic tutor, Paco, for providing a refreshing introduction to classical error correction, for his help in coordinating the administrative details and for connecting me with Miguel Ángel, making this opportunity a reality.
\end{itshape}

\vspace{1cm}
Since many of the people mentioned above do not speak English, I have taken the liberty of including a Spanish translation of these acknowledgments below.

\begin{itshape}

Me gustaría expresar mi más profundo agradecimiento a todos los que han hecho posible este Trabajo de Fin de Máster.

En primer lugar, quiero dar las gracias a mis padres, Frasco y María José, que me inculcaron la libertad, la curiosidad y el valor para cuestionar lo establecido. Mis padres me permitieron crecer libre del peso de las expectativas y siempre valoraron y celebraron mis logros. Estoy igualmente agradecido a mi padrino, Nono, por enseñarme la perseverancia, el amor propio que busca la excelencia y el profundo deseo de dejar el mundo mejor de lo que lo encontré; y a mis abuelos, que me enseñaron el verdadero valor del conocimiento, la sabiduría, la renuncia y el trabajo duro. Y por último, pero no menos importante entre mis raíces, quiero dar las gracias a Lolo, que me enseñó que hay muchas formas distintas, pero igualmente válidas, de entender la vida, y que el camino es más importante que el destino.

Mi mujer, Nines, y nuestros hijos, Pedro, Antonio y Carlos, también merecen toda mi gratitud. No solo son mi principal motivación, sino que han sacrificado tanto como yo por este proyecto. Cursar un Máster pasados los cuarenta, y hacerlo trabajando a tiempo completo, exige inmensos sacrificios familiares, y no lo habría logrado sin su paciencia, comprensión y amor.

Debo mucho a mis profesores de este Máster y, por extensión, a todos mis mentores y maestros, que han construido los cimientos que me han permitido llegar hasta aquí: aquellos que nos enseñan son los arquitectos invisibles de nuestra identidad. Entre ellos, quiero dar las gracias especialmente a Sergio Boixo; una charla suya fue lo que me inspiró por primera vez a adentrarme en el mundo de la computación cuántica, y sus respuestas cuando le pedí consejo me ayudaron a dar mis primeros pasos en este campo. También debo toda mi gratitud a Greg Quillard, un buen amigo que resulta ser la persona que más ha impulsado mi crecimiento profesional a lo largo de mi carrera.

Por último, pero sin duda lo más importante, me gustaría expresar mi sincero agradecimiento al director científico de este trabajo, Miguel Ángel. Ha sido un privilegio absoluto que esta tesis haya sido dirigida por una eminencia en la materia, y es un honor único escribir sobre los códigos de color 2D bajo la tutela de su cocreador. Estoy profundamente agradecido por su tiempo, su visión y por esta extraordinaria oportunidad. Además, hago extensivo mi agradecimiento a mi tutor académico, Paco, por brindarme una refrescante introducción a la corrección de errores clásica, por su ayuda en la coordinación de los trámites administrativos y por ponerme en contacto con Miguel Ángel, haciendo realidad esta oportunidad.
\end{itshape}
